---
title: "01-Agreement-Target-Resolution"
author: "Olena Tykhostup"
date: "2023-12-19"
output: html_document
---
This is the code used for data analysis reported in Chapter 4 of the dissertation "Multimodal annotation and analysis of a Narrative problem solving task in English, Idi, and Russian". This script calculates and visualizes the agreement between three annotators about the targets of pointing gestures annotated in three different conditions: (i) Gesture, card, speech; (ii) Gesture and card; (iii) Gesture only. 

## Load packages
```{r, warning=F, message=F}
library(here)
library(tidyverse)
library(DescTools) # Kappa values CI
library(irr) # Kappa values
```
## Load in the data
```{r}
# Targets in Gesture, card, speech annotation condition (GCSC)
targets_GCSC <- read.csv(here("data", "targets_GestureCardSpeechCondition.csv"))

# Targets in Gesture and card annotation condition (GCC)
targets_GCC <- read.csv(here("data", "targets_GestureCardCondition.csv"))

# Targets in Gesture only annotation condition (GOC)
targets_GOC <- read.csv(here("data", "targets_GestureOnlyCondition.csv"))
```
## Inspect the data
```{r}
# Structure of each dataset
str(targets_GCSC)
str(targets_GCC)
str(targets_GOC)
```
```{r}
# Check for missing values
datasets_list <- list(targets_GCSC, targets_GCC, targets_GOC)

for (i in seq_along(datasets_list)) {
  incomplete_cases <- datasets_list[[i]][!complete.cases(datasets_list[[i]]), ]
  print(incomplete_cases)
}
```

```{r}
# Frequency tables for each dataset
table(stack(targets_GCSC[-1]))
table(stack(targets_GCC[-1]))
table(stack(targets_GOC))
```

## Inspect Fleiss Kappa values
Calculate Fleiss Kappa values for each dataset.
```{r Fleiss}
targets_GCCS_kappa <- kappam.fleiss(targets_GCSC[c(2:4)], 
                                    detail = TRUE)
targets_GCC_kappa <- kappam.fleiss(targets_GCC[c(2:4)], 
                                   detail = TRUE)
targets_GOC_kappa <- kappam.fleiss(targets_GOC, 
                                   detail = TRUE)
```
Construct Confidence intervals for each Kappa point estimate and the overlaps. The Confidence intervals define a range in which the true coefficient lies with a given probability (95%).
```{r Kappa with CI}
# Gesture, card, speech condition (GCSC)
DescTools::KappaM(targets_GCSC[c(2:4)], 
                  method = "Fleiss",
                  conf.level = 0.95)

# Gesture and card condition (GCC)
DescTools::KappaM(targets_GCC[c(2:4)],
                  method = "Fleiss", 
                  conf.level = 0.95)

# Gesture only condition (GOC)
DescTools::KappaM(targets_GOC, 
                  method = "Fleiss",
                  conf.level = 0.95)
```
## Plot Kappa values for each label in each annotation condition
Prepare input data for the plot.
```{r}
# Convert 'targets_GCC_kappa' into a data frame, add a 'Tag' column with annotation labels, 
# and assign a 'Condition' column for Gesture and card annotation condition (GCC)
targets_GCC_kappa.levels <- cbind(as.data.frame.matrix(targets_GCC_kappa$detail), 
                                  Tag = c("card", "interior", "other", "sg")) %>%
  mutate(Condition = "GestureCard")

  
# Data frame with annotation condition Gesture, card, speech (GCSC)
targets_GCCS_kappa.levels <- cbind(as.data.frame.matrix(targets_GCCS_kappa$detail), 
                                   Tag = c("card", "interior", "other", "sg")) %>%
  mutate(Condition = "GestureCardSpeech")


# Data frame with annotation condition Gesture only (GOC)
targets_GOC_kappa.levels <- cbind(as.data.frame.matrix(targets_GOC_kappa$detail), 
                                  Tag = c("card", "interior", "sg")) %>%
  mutate(Condition = "GestureOnly")

# Combine three data frames by row and format for plotting
all_targets_kappa <- rbind(targets_GCC_kappa.levels, 
                           targets_GCCS_kappa.levels, 
                           targets_GOC_kappa.levels) %>%
  mutate_if(is.character, as.factor) %>%
  select(Tag, everything())

# Remove row names
rownames(all_targets_kappa) <- NULL

# Print data frame
all_targets_kappa
```
Generate the plot.
```{r, warning=F, message=F}
# Prepare plot labels
cond.labs <- c(
  paste("Gesture and card \n (n=", nrow(targets_GCC), ")", sep=""),
  paste("Gesture, card and speech \n (n=", nrow(targets_GCSC), ")", sep=""),
  paste("Gesture only \n (n=", nrow(targets_GOC), ")", sep="")
)
names(cond.labs) <- c("GestureCard", 
                      "GestureCardSpeech", 
                      "GestureOnly")
tick_labels <- c("Card", 
                 "Interior", 
                 "Individual", 
                 "Other")

# Set the order of annotation labels
all_targets_kappa$Tag <- factor(all_targets_kappa$Tag,
                                levels = c("card", 
                                           "interior", 
                                           "sg", 
                                           "other"))
# Set the order of plot sections
all_targets_kappa$Condition <- factor(all_targets_kappa$Condition, 
                                      levels = c("GestureCard", 
                                                 "GestureCardSpeech", 
                                                 "GestureOnly")) 

# Plot
targets.kappa.plot <- ggplot(all_targets_kappa %>%
         arrange(Kappa, Tag, Condition), aes(Tag, Kappa)) +
  geom_point(aes(color = Tag), size = 3) +
  geom_segment(aes(x = Tag, xend = Tag, 
                   y = 0, yend = Kappa, 
                   color= Tag), 
               size= 0.7, 
               alpha=0.8) +
  facet_grid(.~ Condition, scales="free", space = "free",
             labeller = labeller(Condition=cond.labs)) + 
  geom_hline(yintercept=0, 
             linetype="dashed", 
             color = "grey40", 
             size = 0.5) +
  scale_x_discrete(labels= tick_labels, expand=c(0.4, 0.3)) +
  theme_minimal() +
  theme(legend.position = "none", 
        strip.text = element_text(size = 12),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        strip.background = element_rect(fill="grey95"),
        axis.title.x = element_text(vjust=-0.9)) + 
  labs(y = "Fleiss' Kappa",
       x = "Annotation label") +
  scale_y_continuous(limits = c(-0.1, 1), breaks = seq(0, 1, by = 0.2)) +
  scale_color_manual(values=c("#183A5A", "#EFB758", "#C34129", "grey10")) +
  geom_rect(aes(xmin=-Inf,xmax=Inf,
                ymin=0.8,ymax=1),
            alpha=0.1,fill="grey60") +
  geom_rect(aes(xmin=-Inf,xmax=Inf,
                ymin=0.6,ymax=0.8),
            alpha=0.1,fill="grey70") +
  geom_rect(aes(xmin=-Inf,xmax=Inf,
                ymin=0.4,ymax=0.6),
            alpha=0.1,fill="grey80") +
  geom_rect(aes(xmin=-Inf,xmax=Inf,
                ymin=0.2,ymax=0.4),
            alpha=0.1,fill="grey90") +
  geom_rect(aes(xmin=-Inf,xmax=Inf,
                ymin=0,ymax=0.2),
            alpha=0.1,fill="grey95")

targets.kappa.plot
```
## Save plot
```{r, warning=FALSE, message=FALSE, echo=FALSE}
# As PNG
png(here("figures", "fppt-three-annotators-targets-kappa.png"), width = 10, height = 6, units = "cm", res = 500)
targets.kappa.plot
dev.off()

# As PDF
pdf(file = here("figures", "fppt-three-annotators-targets-kappa.pdf"), width = 10, height = 6)
targets.kappa.plot
dev.off()
```
## Session info
```{r, warning=F}
sessionInfo()
```
