---
title: "02-Target-Resolution-No-Speech"
author: "Olena Tykhostup"
date: "2024-01-10"
output: 
  html_document:
    fig.path: "Ch4-Pointing_gesture_and_target_resolution/figures/"
---
This is the code used for data analysis reported in Chapter 4 of the dissertation titled 'Multimodal Annotation and Analysis of a Narrative Problem-Solving Task in English, Idi, and Russian'. This script generates mosaic plots to visualize the frequencies of occurrence of four form-related variables and target types as classified by the annotators under the Gesture and Card condition, where gestures were coded without access to concurrent speech. Subsequently, chi-squared tests are conducted, followed by the fitting and visualization of a conditional inference tree. The code also develops a Conditional Random Forest model and evaluates it using Partial Dependence Plots and the visualization of predicted probabilities.

## Load packages
```{r, warning=F, message=F}
library(here)
library(tidyverse)
library(ggmosaic)
library(patchwork)
library(party)
library(partykit)
library(ggparty)
library(caret)
library(pdp)
library(scales)
```
## Define colour palette
```{r}
customColors <- c("#183A5A", "#EFB758", "#C34129", "#E2DBC9")
```
## Load the data
```{r}
noSpeechAnnotations <- read.csv(here("data", "FPPT-EIR-NoSpeech-Annotations.csv"))
```
## Prepare the df for analysis
```{r}
# Select relevant columns, filter data, and ensure no missing values
noSpeechAnnotationsCleaned <- noSpeechAnnotations %>%
  select(speaker, 
         shape, 
         effort, 
         traj, 
         orient, 
         target, 
         lang) %>% 
  filter(!target=="other") %>% 
  mutate_if(is.character,as.factor) %>% 
  drop_na()
```
Inspect the data.
```{r}
# Structure of the cleaned dataframe
str(noSpeechAnnotationsCleaned)
# Check for missing values
sum(is.na(noSpeechAnnotationsCleaned))
# Summary statistics
summary(noSpeechAnnotationsCleaned)
```
## Plot the data
Baloon plot to inspect variation across speakers.
```{r}
# Prepare the data frame for visualization
noSpeechAnnotationsLong <- noSpeechAnnotationsCleaned %>%
  pivot_longer(cols = c(shape, effort, traj, orient), 
               names_to = "category", 
               values_to = "value") %>%
  group_by(speaker, category, value) %>%
  summarise(Freq = n(), .groups = 'drop') %>%
  mutate(category = factor(category, 
                           levels = c("effort", "traj", "orient", "shape"),
                           labels = c("Effort", "Trajectory", "Orientation", "Hand shape")))

# Create a balloon plot with a custom gradient
balloon_plot <- ggplot(noSpeechAnnotationsLong, 
                       aes(x = value, y = speaker, size = Freq, fill = Freq)) +
  geom_point(shape = 21) +
  scale_size(range = c(1, 10)) +
  scale_fill_gradientn(colors = customColors) + 
  facet_wrap(~category, scales = 'free_x', ncol = 4) +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1),
        strip.background = element_blank(),
        strip.text.x = element_text(size = 10)) +
  labs(x = NULL, y = "Speaker", fill = "Frequency")

# Print the plot
balloon_plot
```
## Save plot
```{r, warning=FALSE, message=FALSE, echo=FALSE}
# As PNG
png(here("figures", "fppt-forms-participants-balloon-plot.png"), 
    width = 20, 
    height = 20, 
    units = "cm", 
    res = 500)
balloon_plot
dev.off()

# As PDF
pdf(here("figures", "fppt-forms-participants-balloon-plot.pdf"), 
    width = 6, 
    height = 6)
balloon_plot
dev.off()
```
Mosaic plots for form-related varibales and target types.
```{r, warning=FALSE}
# Change order of factor levels for 'target' column and recode the value 'individual'
levels(noSpeechAnnotationsCleaned$target) <- list(card = "card",
                                                  interior = "interior", 
                                                  individual  = "individ")

# Function definition: geom_mosaicsettings
# This function customizes the appearance of the mosaic plots.
# It adds labels, sets fill colors, applies minimal theme, and adjusts various theme elements.
geom_mosaicsettings <- function() {
    list(
        # Add labels to the plot only if the weight is greater than 4
        geom_label(data = layer_data(last_plot(), 1) %>% 
                   filter(.wt > 4),
                   aes(x = (xmin + xmax) / 2,
                       y = (ymin + ymax) / 2,
                       label = .wt, alpha=0.8)),
        scale_fill_manual(values=customColors),
        theme_minimal(), 
        theme(text = element_text(size=14),
              plot.title = element_text(hjust = 0.5, size = 12),
              axis.text = element_text(size=13),
              axis.text.y = element_text(angle = 90, hjust=0.5),
              legend.position = "none"),
        labs(x = '', y = '')
    )
}

# Create mosaic plot for hand shapes
mosaic.hs <- ggplot(data = noSpeechAnnotationsCleaned) +
  geom_mosaic(aes(x = product(shape, target), fill=shape), color="grey20", size=0.3) +
  labs(title = "Shapes") +
  geom_mosaicsettings()

# Create mosaic plot for trajectories of movement
mosaic.tr <- ggplot(data = noSpeechAnnotationsCleaned) +
  geom_mosaic(aes(x = product(traj, target), fill=traj), color="grey20", size=0.3) +
  labs(title = "Trajectories") +
  geom_mosaicsettings()

# Create mosaic plot for hand orientations
mosaic.or <- ggplot(data = noSpeechAnnotationsCleaned) +
  geom_mosaic(aes(x = product(orient, target), fill=orient), color="grey20", size=0.3) +
  labs(title = "Orientations") +
  geom_mosaicsettings()

# Create mosaic plot for effort
mosaic.ef <- ggplot(data = noSpeechAnnotationsCleaned) +
  geom_mosaic(aes(x = product(effort, target), fill=effort), color="grey20", size=0.3)+
  labs(title = "Effort") +
  geom_mosaicsettings()

# Arrange the above four plots in a 2x2 grid and add a caption at the bottom
complete_mosaic_plot <- (mosaic.hs | mosaic.tr) / (mosaic.or | mosaic.ef) + 
  plot_annotation(caption = 'Targets', 
                  theme = theme(plot.caption = element_text(size = 12, 
                                                            hjust = 0.5)))
complete_mosaic_plot
```
## Save plot
```{r, warning=FALSE, message=FALSE, echo=FALSE}
# As PNG
png(here("figures", "fppt-mosaic-forms-per-target.png"), 
    width = 20, 
    height = 20, 
    units = "cm", 
    res = 500)
complete_mosaic_plot
dev.off()

# As PDF
pdf(here("figures", "fppt-mosaic-forms-per-target.pdf"), 
    width = 10, 
    height = 10)
complete_mosaic_plot
dev.off()
```
## Pearson's Chi-squared test for targets and each poinitng form variable
```{r, warning=F, message=F}
# Get expected frequencies for target and each of the form-related variables for chi-squared tests
noSpeechAnnotationsExpectedFreq <- lapply(noSpeechAnnotationsCleaned[, 2:5], function(x) {
  ch <- chisq.test(noSpeechAnnotationsCleaned$target, x)
  ch$expected
})

noSpeechAnnotationsExpectedFreq

# Perform a series of chi-squared tests on target and each of the form-related variables
noSpeechAnnotationsChi <- t(cbind(apply(noSpeechAnnotationsCleaned[, 2:5], 2, function(x) {
  # Perform chi-squared test
  ch_test <- chisq.test(noSpeechAnnotationsCleaned$target, x)

  # Calculate Cramér's V
  n <- sum(ch_test$observed)  # total sample size
  k <- min(dim(ch_test$observed))  # number of rows or columns, whichever is smaller
  cramers_v <- sqrt(ch_test$statistic / (n * (k - 1)))

  # Combine chi-squared statistic, degrees of freedom, p-value, and Cramér's V
  c(ch_test$statistic, ch_test$parameter, p_value = ch_test$p.value, cramers_v=cramers_v)
})))

# Print the results
print(noSpeechAnnotationsChi)
```
## Conditional inference trees for target types
```{r, warning=FALSE}
# Set a random seed for reproducibility
set.seed(60)

# Fit a Conditional Inference Tree (ctree) for target types
gtree.target <- partykit::ctree(target ~ shape + orient + traj + effort,
                      data = noSpeechAnnotationsCleaned)

# Plot the conditional inference tree
plot(gtree.target)

# Fit another ctree without 'effort' for cleaner visualization
gtree.target_noeffort <- partykit::ctree(target ~ shape + orient + traj,
                       data = noSpeechAnnotationsCleaned)

# Create a ggparty plot for the conditional inference tree
ggtree.target.plot <- ggparty(gtree.target_noeffort, terminal_space = 0.6,
                              add_vars = list(p.value = "$node$info$p.value")) +
  geom_edge() +
  geom_edge_label(col = "grey24", size = 4) +
  geom_node_plot(gglist = list(geom_bar(aes(x = "", fill = target), 
                                      width = 0.6, alpha = 0.8, 
                                      colour = "grey20", size = 0.5, 
                                      position = position_fill()),
                               scale_fill_manual(values = customColors),
                               scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                                                  breaks = NULL),
                               labs(fill = "Targets"),
                               xlab(""), ylab("Pointing gestures"),
                               theme_minimal(),
                               theme(text = element_text(size = 12)),
                               geom_label(aes(x = "", group = target, 
                                              label = ifelse(..count.. > 3, ..count.., "")),
                                          stat = "count", 
                                          position = position_fill(), 
                                          vjust = 0.9, alpha = 0.8,
                                          size = 3)),
                 shared_axis_labels = TRUE) + 
  geom_node_label(aes(),
                  line_list = list(aes(label = paste("Node", id)),
                                   aes(label = splitvar),
                                   aes(label = paste(scales::pvalue(p.value, 
                                                         accuracy = 0.001,
                                                         decimal.mark = ".",
                                                         add_p = TRUE)))),
                  line_gpar = list(list(size = 8, col = "black"),
                                   list(size = 10, fontface = "bold"),
                                   list(size = 8)),
                  ids = "inner") 

# Print the ggparty plot
ggtree.target.plot
```
## Save plot
```{r, warning=FALSE, message=FALSE, echo=FALSE}
# As PNG
ggsave(here("figures", "ctree-target-form.png"), 
       plot = ggtree.target.plot, 
       width = 10, 
       height = 8, 
       dpi = 300)

# As PDF 
ggsave(here("figures", "ctree-target-form.pdf"), 
       plot = ggtree.target.plot, 
       width = 10, 
       height = 8)
```
## How well does the tree represent the data?
```{r}
# Create a confusion matrix for the tree's predictions
tree_predictions <- predict(gtree.target, type = "response")

# Generate a confusion matrix to evaluate model performance
confusion_matrix <- confusionMatrix(tree_predictions, 
                                    noSpeechAnnotationsCleaned$target)

# Print the confusion matrix to assess model representation of the data
confusion_matrix
```
The accuracy is 0.6643 means that the correct predictions are made for 66% of observations.

## Tree with language as predictor
```{r, warning=FALSE}
set.seed(60)
gtree.target.lang <- partykit::ctree(target ~ shape + orient + traj + lang,
                          data = noSpeechAnnotationsCleaned)

ggtree.target.lang.plot <- ggparty(gtree.target.lang, terminal_space = 0.4,
        add_vars = list(p.value = "$node$info$p.value")) +
  geom_edge() +
  geom_edge_label(col="grey24", size=4) +
  geom_node_plot(gglist = list(geom_bar(aes(x = "", fill = target), 
                                        width = 0.8, alpha = 0.8, 
                                        colour="grey20", size = 0.3, 
                                        position = position_fill()),
                               scale_fill_manual(values = customColors),
                               scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                                                  breaks=NULL),
                               labs(fill = "Targets"),
                               xlab(""), ylab("Pointing gestures"),
                               theme_minimal(),
                               theme(text = element_text(size=12))),
                 shared_axis_labels = T)  +
  geom_node_label(aes(),
                  line_list = list(aes(label = paste("Node", id)),
                                   aes(label = splitvar),
                                   aes(label = paste(scales::pvalue(p.value, 
                                                                    accuracy = 0.001,
                                                                    decimal.mark = ".",
                                                                    add_p = T)))),
                  line_gpar = list(list(size = 8, col = "black"),
                                   list(size = 10, fontface = "bold"),
                                   list(size = 8)),
                  ids = "inner") +
  # add labels for terminal nodes
  geom_node_label(aes(label = paste0("N = ", nodesize)),
                  ids = "terminal",
                  size = 3,
                  line_gpar = list(list(size = 8, col = "black"),
                                   list(size = 10, fontface = "bold"))) +
  labs(title = "") +
  geom_rect(aes(xmin=0.01, xmax=0.365, ymin=0.040, ymax=0.418), 
            color = "#EFB758", fill=NA) +
  geom_rect(aes(xmin=0.385, xmax=0.615, ymin=0.040, ymax=0.418), 
            color = "#C34129", fill=NA) +
  geom_rect(aes(xmin=0.635, xmax=0.740, ymin=0.040, ymax=0.418),
            color = "grey", fill=NA) +
  geom_rect(aes(xmin=0.760, xmax=0.990, ymin=0.040, ymax=0.418),
            color = "#183A5A", fill=NA)

# Print plot
ggtree.target.lang.plot
```
## Save plot
```{r, warning=FALSE, message=FALSE, echo=FALSE}
# As PNG
ggsave(here("figures", "ctree-target-form-lang.png"), 
       plot = ggtree.target.lang.plot, 
       width = 10, 
       height = 8, 
       dpi = 300)

# As PDF 
ggsave(here("figures", "ctree-target-form-lang.pdf"), 
       plot = ggtree.target.lang.plot, 
       width = 10, 
       height = 8)
```

## How well does the tree represent the data?
```{r}
# Create a confusion matrix for the tree's predictions
tree_predictions_wlang <- predict(gtree.target.lang, 
                                  type = "response")

# Generate a confusion matrix to evaluate model performance
confusion_matrix_wlang <- confusionMatrix(tree_predictions_wlang, 
                                          noSpeechAnnotationsCleaned$target)

# Print
confusion_matrix_wlang
```
# Random forest model and conditional variable importance scores
```{r, warning=F, message=F}
set.seed(30)
# Fit a Conditional Random Forest model (cforest) for target types
rf.target <- party::cforest(target ~ shape + orient + traj + effort,
                          data = noSpeechAnnotationsCleaned,
                          control=party::cforest_unbiased(mtry=2,ntree=1000))

# Compute the conditional variable importance scores
varimp.target <- round(party::varimp(rf.target, conditional = TRUE), 3)

# Create a chart with conditional importance scores
dotchart(sort(varimp.target), main = "Conditional importance of variables")

# Plot a red line to separate the important scores from the unimportant scores
abline(v=abs(min(varimp.target)), lty=2, lwd=2, col="red")

# Get accuracy
caret::confusionMatrix(data = predict(rf.target), 
                       reference = noSpeechAnnotationsCleaned$target, 
                       mode = "prec_recall")
```
## Plot Random Forest model results as conditional importance plot
```{r, warning=F, message=F}
varimp.df <- as.data.frame(varimp.target)

# Row names to column
varimp.df$names <- rownames(varimp.df)

cond.imp.plot <- ggplot(varimp.df, 
                        aes(x=reorder(names, varimp.target), 
                                       y=varimp.target)) + 
  geom_point(color= "#183A5A", size=4, alpha=0.9) +
  geom_hline(yintercept = 0.005, 
             linetype="dashed", 
             color = "grey40", 
             size = 0.5) +
  geom_segment(aes(x=names, 
                   xend=names, 
                   y=0, 
                   yend=varimp.target), 
               color= "#183A5A", size= 1, alpha=0.8) +
  labs(x="", y="") +
  coord_flip() +
  theme_minimal() +
  geom_vline(xintercept=0.10, 
             linetype="dashed", 
             color = "grey40", 
             size = 0.5) +
  theme(aspect.ratio = 2/3,
        text = element_text(size=12),
        axis.text = element_text(size=12),
        axis.title.x = element_text(vjust=-1.5),
        axis.title.y = element_text(vjust=2)) +
  scale_x_discrete(labels=c("traj" = "Trajectory", 
                            "shape" = "Hand shape",
                            "orient" = "Orientation",
                            "effort" = "Effort"))
cond.imp.plot
```
## Save plot
```{r, warning=FALSE, message=FALSE, echo=FALSE}
# As PNG
ggsave(here("figures", "cond-imporance-plot-target-form.png"), 
       plot = cond.imp.plot, 
       width = 10, 
       height = 8, 
       dpi = 300)

# As PDF 
ggsave(here("figures", "cond-imporance-plot-target-form.pdf"), 
       plot = cond.imp.plot, 
       width = 10, 
       height = 8)
```

## Partial dependecne plots
```{r, warning=F, message=F}
# Define a list of variables and their names for partial dependence plots
variables <- c("traj", "shape", "orient", "effort")
names <- c("Trajectory", "Shape", "Orientation", "Effort")
names_map <- setNames(names, variables)

# Initialize an empty list to store ggplot objects
pdp_plots <- list()

# Loop through each variable to create partial dependence plots 
# and convert to data frames
for (var in variables) {
  # Calculate partial dependence
  pdp <- partial(rf.target, var, prob = TRUE)
  
  # Convert to data frame
  pdp_df <- as.data.frame(pdp)

  # Create a ggplot for each variable
  pdp_plots[[var]] <- ggplot(pdp_df, aes_string(x = var, y = "yhat")) +
    geom_point(color = "#183A5A", size = 4, alpha = 0.9) +  
    geom_segment(aes_string(x = var, xend = var, y = 0, yend = "yhat"), 
                 color = "#183A5A", size = 1, alpha = 0.8) + 
    labs(x = names_map[var], y = "") +  
    theme_minimal() +
    theme(aspect.ratio = 2/3,
          text = element_text(size = 12),
          axis.text = element_text(size = 12),
          axis.title.x = element_text(vjust = -1.5),
          axis.title.y = element_text(vjust = 2))
}

# Combine and arrange the plots using patchwork
complete_pdp_plot <- wrap_plots(pdp_plots, ncol = 2) + 
  plot_annotation(theme = theme(plot.caption = element_text(size = 12, 
                                                            hjust = 0.5)))

# Print the combined plot
complete_pdp_plot
```
## Save plot
```{r, warning=FALSE, message=FALSE, echo=FALSE}
# As PNG
ggsave(here("figures", "part-dep-plots-target-forms.png"), 
       plot = complete_pdp_plot, 
       width = 20, 
       height = 20, 
       units = "cm")

# Save as PDF
ggsave(here("figures", "part-dep-plots-target-forms.pdf"), 
       plot = complete_pdp_plot, 
       width = 20, 
       height = 20)
```
# Predict and visualize target type probabilities using the Random Forest model
```{r, warning=F, message=F}
# Generate probabilities using the random forest model
probabilities <- predict(rf.target, 
                         newdata = noSpeechAnnotationsCleaned, 
                         type = "prob")

# Combine the probabilities into a data frame
prob_df <- do.call(rbind, probabilities)

# Reshape the data frame for visualization
long_prob_df <- as.data.frame(prob_df) %>%
  mutate(actual_class = noSpeechAnnotationsCleaned$target) %>%
  pivot_longer(cols = starts_with("target"), 
               names_to = "predicted_class", 
               values_to = "probability") %>%
  mutate(predicted_class = str_replace(predicted_class, "target.", "")) %>%
  mutate(predicted_class=as.factor(predicted_class)) %>%
  mutate(
    actual_class = fct_recode(actual_class, "individual" = "individ"),
    predicted_class = fct_recode(predicted_class, "individual" = "individ")
  ) %>%
  mutate(
    actual_class = fct_relevel(actual_class, "card", "interior", "individ"),
    predicted_class = fct_relevel(predicted_class,"card", "interior", "individ")
  )

# Create a boxplot to visualize predicted probabilities and the actual targets
pred.prob.plot <- ggplot(long_prob_df, aes(x = actual_class, 
                                           y = probability, 
                                           fill = predicted_class)) +
  geom_boxplot(alpha=0.8, color="grey20", size=0.3) +
  labs(x = "Actual Target",
       y = "Predicted Probability",
       fill = "Predicted Targets") +
  theme_minimal() +
  scale_fill_manual(values = customColors) +
  theme(legend.position = "bottom",
        text = element_text(size = 12),
        axis.text = element_text(size = 12),
        axis.title.x = element_text(vjust = -1.5),
        axis.title.y = element_text(vjust = 2),
        legend.text = element_text(size = 12))

# Print the probability plot
pred.prob.plot
```

## Save plot
```{r, warning=FALSE, message=FALSE, echo=FALSE}
# As PNG
ggsave(here("figures", "targets-pred-prob-plot.png"), 
       plot = pred.prob.plot, 
       width = 20, 
       height = 15, 
       units = "cm")

# Save as PDF
ggsave(here("figures", "targets-pred-prob-plot.pdf"), 
       plot = pred.prob.plot, 
       width = 10, 
       height = 8)
```
## Session info
```{r, warning=F}
sessionInfo()
```